---
title: "Rational Numbers: An example of R Classes"
author: "Rob Carnell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rational Numbers: An example of R Classes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
require(rational)
require(R6)
```

## Why do we need a rational package?

This package serves 2 purposes:

- Demonstrates the creation of the same classes in S3, S4, and R6 class systems
- Demonstrates one way to solve some numerical accuracy problems that people struggle with

For example:

```{r demo}
# expectations dashed
(0.1 + 0.2) == 0.3
# what?
print(0.1 + 0.2, digits = 20)
# what am I supposed to do in R? (or any other floating point arithmetic system)
all.equal(0.1 + 0.2, 0.3, tolerance = 1E-9)
abs(0.1 + 0.2 - 0.3) < 1E-9
# is there another way?
# Yes, rational numbers
#   NOTE: the "L" notation indicates an integer
rational(1L, 10L) + rational(2L, 10L) == rational(3L, 10L)
```

## Initialization

#### Rational S3

```{r rational_s3_init, eval=FALSE}
.rationalS3 <- function(n, d)
{
  ret <- list(n = n, d = d, v = n / d)
  class(ret) <- "rationalS3"
  return(ret)
}
```

```{r rational_S3_init2}
# generating function
a <- rational(1L, 3L, method = "S3")
# basic structure
str(a)
# what is this object?
class(a)
is.list(a)
is.rational(a)
is.rationalS3(a)
is.numeric(a)
is.integer(a)
# how can I access the values?
a$n
a$d
a$v
```

#### Rational S4

```{r rational_s4_init, eval = FALSE}
setClass("rationalS4", slots = c(n = "integer", d = "integer", v = "numeric"),
         valid = function(object)
         {
           if (length(object@n) == length(object@d)) {
             if (all(is.integer(object@n)) && all(is.integer(object@d))) {
               if (!any(object@d == 0)) return(TRUE)
               else return(.rationalErrorMessage2)
             }
             else return(.rationalErrorMessage1)
           }
           else return(.rationalErrorMessage3)
         })

setMethod("initialize", "rationalS4", function(.Object, n, d)
{
  .Object@n <- n
  .Object@d <- d
  .Object@v <- n / d
  # validity checks happen on the default initialize
  callNextMethod(.Object = .Object, n = n, d = d)
})
```

```{r rational_s4_init2}
# generating function
a <- rational(1L, 3L, method = "S4")
# basic structure
str(a)
# what is this object?
class(a)
is.rational(a)
is.rationalS4(a)
is.numeric(a)
is.integer(a)
# how can I access the values?
a@n
a@d
a@v
```

#### Rational R6

```{r rational_R6_init, eval=FALSE}
.rationalR6 <- R6Class("rationalR6",
  public = list(
   getNumerator = function() private$n,
   getDenominator = function() private$d,
   getValue = function() private$v,
   initialize = function(n, d)
   {
     private$n <- n
     private$d <- d
     private$v <- n / d
     self
   },
   setNumerator = function(x)
   {
     private$n <- x
     private$v <- private$n / private$d
   },
   setDenominator = function(x)
   {
     private$d <- x
     private$v <- private$n / private$d
   },
   assign_at = function(i, value)
   {
     private$n[i] <- value$getNumerator()
     private$d[i] <- value$getDenominator()
     private$v <- private$n / private$d
   }),
  private = list(
   n = 1L,
   d = 1L,
   v = 1L
  ), lock_class = FALSE, lock_objects = TRUE, portable = TRUE)
```

```{r rational_R6_init2}
# generating function
a <- rational(1L, 3L, method = "R6")
# basic structure
str(a)
# what is this object?
class(a)
is.rational(a)
is.rationalR6(a)
is.numeric(a)
is.integer(a)
# how can I access the values?
a$getNumerator()
a$getDenominator()
a$getValue()
```

## Create operators

#### Addition

```{r add, eval=FALSE}
'+.rationalS3' <- function(e1, e2)
{
  if (is.rationalS3(e1) && is.rationalS3(e2))
  {
    res <- .rationalAddRational(e1$n, e1$d, e2$n, e2$d)
    return(.rationalS3(res$n, res$d))
  } else if (is.integer(e1) && is.rationalS3(e2))
  {
    res <- .rationalAddInteger(e2$n, e2$d, e1)
    return(.rationalS3(res$n, res$d))
  } else if (is.rationalS3(e1) && is.integer(e2))
  {
    res <- .rationalAddInteger(e1$n, e1$d, e2)
    return(.rationalS3(res$n, res$d))
  } else if (is.numeric(e1) && is.rationalS3(e2))
  {
    return(.rationalAddNumeric(e2$n, e2$d, e1))
  } else if (is.rationalS3(e1) && is.numeric(e2))
  {
    return(.rationalAddNumeric(e1$n, e1$d, e2))
  } else
  {
    return(NA)
  }
}

'+.rationalR6' <- function(e1, e2)
{
  if (is.rationalR6(e1) && is.rationalR6(e2))
  {
    res <- .rationalAddRational(e1$getNumerator(), e1$getDenominator(), 
                                e2$getNumerator(), e2$getDenominator())
    return(.rationalR6$new(res$n, res$d))
  } else if (is.integer(e1) && is.rationalR6(e2))
  {
    res <- .rationalAddInteger(e2$getNumerator(), e2$getDenominator(), e1)
    return(.rationalR6$new(res$n, res$d))
  } else if (is.rationalR6(e1) && is.integer(e2))
  {
    res <- .rationalAddInteger(e1$getNumerator(), e1$getDenominator(), e2)
    return(.rationalR6$new(res$n, res$d))
  } else if (is.numeric(e1) && is.rationalR6(e2))
  {
    return(.rationalAddNumeric(e2$getNumerator(), e2$getDenominator(), e1))
  } else if (is.rationalR6(e1) && is.numeric(e2))
  {
    return(.rationalAddNumeric(e1$getNumerator(), e1$getDenominator(), e2))
  } else
  {
    return(NA)
  }
}

setMethod("+", c("rationalS4", "rationalS4"), function(e1, e2)
{
  res <- .rationalAddRational(e1@n, e1@d, e2@n, e2@d)
  return(new("rationalS4", n = res$n, d = res$d))
})

setMethod("+", c("integer", "rationalS4"), function(e1, e2)
{
  res <- .rationalAddInteger(e2@n, e2@d, e1)
  return(new("rationalS4", n = res$n, d = res$d))
})

# and many more ...
```

```{r add_show}
a3 <- rational(1L, 3L, method = "S3")
b3 <- rational(3L, 4L, method = "S3")
a3 + 1.8
a3 + 2L
a3 + b3
```

#### Other Operators

```{r log10, eval=FALSE}
.rational_log <- function(n, d, base)
{
  if (base == exp(1))
    log(n) - log(d)
  else if (base == 10)
    log10(n) - log10(d)
  else if (base == 2)
    log2(n) - log2(d)
  else
    logb(n, base = base) - logb(d, base = base)
}

setMethod("log10", signature = c("rationalS4"),
  function(x)
  {
    .rational_log(x@n, x@d, 10)
  }
)

log10.rationalS3 <- function(x)
{
  .rational_log(x$n, x$d, 10)
}

log10.rationalR6 <- function(x)
{
  .rational_log(x$getNumerator(), x$getDenominator(), 10)
}
```

```{r log10_show}
log10(rational(1L, 3L, method = "S3"))
log10(rational(3L, 4L, method = "R6"))
log10(rational(3L, 4L, method = "S4"))
```

## Inheritance

#### S3

```{r s3inherit}
polygon <- function(area)
{
  value <- list(area = area)
  class(value) <- "polygonS3"
  return(value)
}

rectangle <- function(l, w)
{
  value <- list(area = l*w, l = l, w = w)
  class(value) <- c("rectangleS3", "polygonS3")
  return(value)
}

print.polygonS3 <- function(obj) cat("Area: ", obj$area, "\n")
print.rectangleS3 <- function(obj) 
{
  cat("Length: ", obj$l, " Width: ", obj$w, " ")
  print.polygonS3(obj)
}

p3 <- polygon(5)
r3 <- rectangle(2, 3)
is(p3, "polygonS3")
is(r3, "polygonS3")
is(r3, "rectangleS3")
inherits(r3, "polygonS3")
p3
r3
```

#### S4

```{r s4inherit}
setClass("polygonS4", slots = list(area = "numeric"))
setMethod("show", "polygonS4", function(object) cat("Area: ", object@area, "\n"))
setClass("rectangleS4", slots = list(l = "numeric", w = "numeric"),
         contains = "polygonS4")
setMethod("initialize", "rectangleS4", function(.Object, l, w)
{
  .Object@l <- l
  .Object@w <- w
  .Object@area <- l*w
  .Object
})
setMethod("show", "rectangleS4", function(object) 
{
  cat("Length: ", object@l, " Width: ", object@w, " ")
  callNextMethod()
})
p4 <- new("polygonS4", area = 5)
r4 <- new("rectangleS4", l = 2, w = 3)
is(p4, "polygonS4")
is(r4, "polygonS4")
is(r4, "rectangleS4")
inherits(r4, "polygonS4")
p4
r4
```

#### R6

```{r r6inherit}
polygonR6 <- R6Class("polygonR6",
  public = list(
    initialize = function(area)
    {
      private$area = area
    },
    print = function()
    {
      cat("Area: ", private$area, "\n")
    }
  ),
  private = list(
    area = numeric()
  )
)

rectangleR6 <- R6Class("rectangleR6",
  inherit = polygonR6,
  public = list(
    initialize = function(l, w)
    {
      private$l = l
      private$w = w
      private$area = l*w
    },
    print = function()
    {
      cat("Length: ", private$l, " Width: ", private$w, " ")
      super$print()
    }
  ),
  private = list(
    l = numeric(),
    w = numeric()
  )
)
p6 <- polygonR6$new(area = 5)
r6 <- rectangleR6$new(l = 2, w = 3)
is(p6, "polygonR6")
is(r6, "polygonR6")
is(r6, "rectangleR6")
inherits(r6, "polygonR6")
p6
r6
```

## Encapsulation

#### S3 + S4

There is no encapsulation in the S3 and S4 system.  All objects of the class can be 
accessed directly.  There is no concept of "public" and "private" in the class.

```{r encaps3}
r3$l <- 10
# this is bad
r3

r4@l <- 10
# so is this
r4

# R6 solves this problem
tryCatch(r6$l <- 10, error = function(e) print(e)) 
```

#### R6

With R6, we can restrict the method users use to interact with the class.

```{r encapr6}
rectangleR6$set("public", "setLength", function(l){
  private$l <- l
  private$area <- l * private$w
})
r6 <- rectangleR6$new(2, 3)
r6$setLength(10)
r6

```

